### 2019-08-29 concepts.tex
changes to formalism/intro largely done

{\bf If a call fails, and indicates this to the caller, it should not modify system
  state in any way.  Changes made before the failure should be rolled
  back.}

@@

In languages with a clear mechanism for expressing failure of a call
(e.g., exceptions in Python and Java, or Option/Result types in Rust,
Haskell, and ML), failure determinism can be automatically checked.
Moreover, the expected overhead should be even lower than the general
overhead for vertical determinism, in that we can expect most failing
operations to be fast (since in test generation, failure is usually
due to invalid parameters).  Checking equivalence of the full
observable state, however, is still expensive, and requires defining
the observable components of a state.  A more lightweight, but (as we
will show) still quite powerful way of checking for failure nondeteterminism is to
repeat the failing operation, and check if it still fails.  If not,
this is a sign that the operation is not deterministic with respect to
failure.  Note that this more restrictive notion of failure
nondeterminism captures such issues as the Apple login bug.

@@

@@

Vertical nondeterminism, in contrast, is demonstrated by a single trace
$t$, not a pair of traces.  To define vertical nondeterminism with
respect to idempotency, we can
define $\mathcal{I} \subseteq A$, the set of \emph{idempotent}
actions.  A trace $t s_0
\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} s_n$ shows \emph{vertical nondeterminism
  with respect to idempotent operations} if $\exists i < n$ such that
$\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_{i+1})
\neq v(s_{i+2})$.  Note that vertical nondeterminism may exhibit only
after a sequence of more than two successive applications of a (supposedly)
idempotent operation; the definition only requires that after \emph{some}
sequence of actions, possibly with $\alpha_{i-1} = \alpha_i$, the
visible state changes.

To define failure nondeterminism, we extend the
transition relation to include a notion of \emph{failure}: $T \subseteq S
\times A \times S \times {\tt bool}$, where the boolean indicates
whether the action \emph{failed}.  A trace $t = s_0
\xrightarrow[]{\alpha_0} (s_1, F_0) \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} (s_n, F_n)$ shows \emph{failure
  nondeterminism} if $\exists i$ such that $F_{i+1}$ (that is,
$\alpha_i$ \emph{fails}) and either (1) $v(s_{i}) \neq v(s_{i+1})$
(the visible state changes) or (2)
$\alpha_i = \alpha_{i+1} \wedge \neg F_{i+2}$ (the same action is
repeated, but this time does not fail).  The second possibility
offers a \emph{lightweight} approach to checking for failure
nondeterminism.  Unlike all of the other ways to demonstrate
nondeterminism we consider, this second type of failure nondeterminism does not
require the use of $v$, or even the notion of observable components of
a state, or (for that matter) any comparison of states.  It only
requires that we be able to say whether an operation failed.

A further point to emphasize is that while horizontal determinism
requires two traces, it does not require a test generation algorithm
to produce pairs of traces, which would require substantial
modification of most such tools (e.g., random testers or fuzzers, and
even model checkers).  In general, testing tools do not produce
\emph{traces}, they produce \emph{action sequences}.  The two traces
used to show nondeterminism, by definition, have the same action
sequences, which means that a testing tool only needs to search for an
action sequence which, when executed twice, produces a $t_1$ and $t_2$
meeting the requirements for showing nondeterminism ---
non-equivalence of $v(s)$ for some $s$.

>>>>

{\bf If a call \emph{fails}, and indicates this to the caller, it should not modify system
  state in any way; changes should be ``rolled back.''}

@@

In languages with a clear mechanism for expressing failure of a call
(e.g., exceptions in Python and Java, or Option/Result types in Rust,
Haskell, and ML), failure determinism can be automatically checked,
given a notion of general vertical determinism.
Moreover, the expected overhead should be even lower than the usual
overhead for vertical determinism, in that we can expect most failing
operations to be fast (since in test generation, failure is usually
due to invalid parameters).  Checking equivalence of the full
observable state, however, is still expensive, and requires defining
the observable components of a state.  A more lightweight, but (as we
will show) still quite powerful way of checking for failure nondeteterminism is to
repeat the failing operation, and check if it still fails.  If not,
this is a sign that the operation is not deterministic with respect to
failure.  Note that this more restrictive notion of failure
nondeterminism captures such issues as the Apple login bug.  The
changes to visible or hidden state we are most concerned with are
usually those related to the failure itself.

@@

The concept of an action does not restrict this approach to reactive
systems; it can be generalized to consider the only action that varies
to be the selection of an input, $\alpha_0$, with the remainder of the
actions being internal $\tau$ actions, in the usual fashion.

\subsubsection{Horizontal Determinism}

@@

A point to emphasize is that while horizontal determinism
requires two traces, it does not require a test generation algorithm
to \emph{produce} pairs of traces, which would require substantial
modification of most such tools (e.g., random testers or fuzzers, and
even model checkers).  In general, testing tools do not produce
\emph{traces}, they produce \emph{action sequences} or \emph{inputs}.  The two traces
used to show nondeterminism, by definition, have the same action
sequences, which means that a testing tool only needs to search for an
action sequence or input which, when executed twice, produces a $t_1$ and $t_2$
meeting the requirements for showing nondeterminism ---
non-equivalence of $v(s)$ for some $s$.

\subsubsection{Vertical Determinism}

Vertical nondeterminism, in contrast, is demonstrated by a single trace
$t$, not a pair of traces.  To define vertical nondeterminism with
respect to idempotency, we can
define $\mathcal{I} \subseteq A$, the set of \emph{idempotent}
actions.  A trace $t s_0
\xrightarrow[]{\alpha_0} s_1 \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} s_n$ shows \emph{vertical nondeterminism
  with respect to idempotent operations} if $\exists i < n$ such that
$\alpha_i \in \mathcal{I} \wedge \alpha_i = \alpha_{i+1}$ and $v(s_{i+1})
\neq v(s_{i+2})$.

Note that vertical nondeterminism may, in theory, exhibit only
after a sequence of more than two successive applications of a (supposedly)
idempotent operation; the definition only requires that after \emph{some}
sequence of actions, possibly with $\alpha_{i-1} = \alpha_i$, the
visible state changes.  In an implementation, we would usually restrict the
search to a finite number of repetitions; for many faults, two
``copies'' will suffice, since an operation is either truly idempotent or
not idempotent.  Only the existence of \emph{non-visible} state changes that
can eventually produce visible state changes \emph{after a succession of the
same action} require more repetitions.

To define failure nondeterminism, we extend the
transition relation to include a notion of \emph{failure}: $T \subseteq S
\times A \times S \times {\tt F: bool}$, where the boolean $F$ indicates
whether the action $A$ \emph{failed}.

A trace $t = s_0
\xrightarrow[]{\alpha_0} (s_1, F_0) \xrightarrow[]{\alpha_1} \ldots
\xrightarrow[]{\alpha_{n-1}} (s_n, F_{n-1})$ shows \emph{failure
  nondeterminism} if $\exists i$ such that $F_{i}$ (that is,
$\alpha_i$ \emph{fails}) and either (1) $v(s_{i}) \neq v(s_{i+1})$
(the visible state changes) or (2)
$\alpha_i = \alpha_{i+1} \wedge \neg F_{i+1}$ (the same action is
repeated, but this time does not fail).  The second possibility
offers a \emph{lightweight} approach to checking for failure
nondeterminism.  Unlike all of the other ways to demonstrate
nondeterminism we consider, this second type of failure nondeterminism does not
require the use of $v$, the notion of observable components of
a state, or any comparison of states.  It only
requires that we be able to say whether an operation failed.
